---
title: "Lec 14: Single Table Analysis"
subtitle: "SDS 192: Introduction to Data Science"
author: |
  Shiya Cao<br/>
  <span style = 'font-size: 70%;'>
  [Statistical & Data Sciences](http://www.smith.edu/sds), Smith College<br/>
  </span>
date: |
  Fall 2022<br/>
  `r sds::img_logo(width = 64)`
output:
  sds::moon_reader:
    lib_dir: libs
    css: xaringan-custom.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false    
      navigation:
        scroll: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=300)
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE}
# sds::duo_smith()
sds::mono_light_smith()
```

```{r include=FALSE}
library(tidyverse)
hospitals <- read.csv("https://raw.githubusercontent.com/SDS-192-Intro/SDS-192-public-website/main/slides/datasets/Hospital_Cost_Report_2018.csv")
```

# Today's Learning Goals

* Conduct single table analysis using `arrange()`, `summarize()`, `group_by()`, `mutate()`.

---

# Today's Dataset

* Hospital Cost Reports, 2018

```{r}
library(tidyverse)
hospitals <- read.csv("data/Hospital_Cost_Report_2018.csv")
```

---

# `arrange()`

* `arrange()` sorts rows according to values in a column.
* Defaults to sorting from smallest to largest (numeric) or first character to last character (character).

```{r}
hospitals |>
  filter(City == "SPRINGFIELD" &amp; State.Code == "MA") |>
  arrange(Number.of.Beds) |>
  select(Hospital.Name, Number.of.Beds)
```

---

# `arrange()`

* `desc()` sorts from largest to smallest (numeric) or last character to first character (character).

```{r}
hospitals |>
  filter(City == "SPRINGFIELD" &amp; State.Code == "MA") |>
  arrange(desc(Number.of.Beds)) |>
  select(Hospital.Name, Number.of.Beds)
```

---

# `summarize()`

* `summarize()` computes a value across a vector of values and stores it in a new data frame.


```r
summary_hospitals_ma <- hospitals |>
  filter(State.Code == "MA") |>
  summarize(total_beds = sum(Number.of.Beds))
summary_hospitals_ma
```

```
## # A tibble: 1 × 1
##   total_beds
##        &lt;int&gt;
## 1         NA
```

&gt; Why did we get an `NA` value?

---
      
# Let's filter to MA


```r
hospitals_ma &lt;-
  hospitals |>
  filter(State.Code == "MA")
```
 
---
      
# `summarize()`

* Setting `na.rm` to TRUE ignore any missing values and only returns the summary value for all non-missing values.
* `sum(is.na())` computes the total `NA` values in the vector.


```r
summary_hospitals_ma <- hospitals_ma |>
  summarize(total_beds = sum(Number.of.Beds, na.rm = TRUE), 
            total_beds_na = sum(is.na(Number.of.Beds)))
summary_hospitals_ma
```

```
## # A tibble: 1 × 2
##         total_beds       total_beds_na
##        &lt;int&gt;         &lt;int&gt;
## 1            18804                   2
```

&gt; What if I wanted to know the percentage of rows with `NA` values in the `Number.of.Beds` column?
      
---
      
# `summarize()`

* `n()` computes the total observations in the vector.


```r
summary_hospitals_ma <- hospitals_ma |>
  summarize(total_beds = sum(Number.of.Beds, na.rm = TRUE), 
            percent_beds_na = sum(is.na(Number.of.Beds)) / n() * 100)
summary_hospitals_ma
```

```
## # A tibble: 1 × 2
##         total_beds       percent_beds_na
##        &lt;int&gt;           &lt;dbl&gt;
## 1            18804                  2.04
```

---
 
class: center, middle
# But there are very different types of hospitals represented in this dataset. Maybe I want to know the number of beds for each type of hospital.

---
      
# `group_by()` |> `summarize()`

* `group_by()` groups observations with a shared value in a variable.
* When we combine `group_by()` and `summarize()` we can perform operations *within* groups.


```r
hospitals_ma |>
  group_by(CCN.Facility.Type) |>
  summarize(total_beds = sum(Number.of.Beds, na.rm = TRUE), 
            total_hopsitals = n(),
            percent_beds_na = sum(is.na(Number.of.Beds)) / n() * 100)
```

---

```
## `summarize()` has grouped output by 'CCN.Facility.Type'.
```
      
```
## # A tibble: 7 × 4
##   CCN.Facility.Type total_beds total_hopsitals percent_beds_na
##   &lt;chr&gt;      &lt;int&gt;     &lt;int&gt;     &lt;dbl&gt;
## 1 CAH                       74               3               0
## 2 CH                       527               4              50
## 3 LTCH                    2416              11               0
## 4 PH                      1704              17               0
## 5 RH                       752               6               0
## 6 RNMHC                     20               1               0
## 7 STH                    13311              56               0
```

---
      
# `group_by()`

* Grouping only changes the metadata of a data frame.
* `ungroup()` removes the grouping structure metadata.


```r
hospitals_ma |>
  group_by(CCN.Facility.Type) |>
  ungroup()
```
  
---
      
# `mutate()`

* `mutate()` creates a new variable (column) in a data frame and fills values according to criteria we provide.

&gt; What would the following lines of code do?


```r
hospitals_ma |>
  mutate(Address = paste(Street.Address, City, State.Code, Zip.Code))

hospitals_ma |>
  mutate(Hospital.Name.Lower = tolower(Hospital.Name))

hospitals_ma |>
  mutate(Percentage.State.Beds = Number.of.Beds/sum(Number.of.Beds, na.rm = TRUE) * 100)
```
---
      
# `mutate()`


```r
hospitals_ma |>
  mutate(Address = paste(Street.Address, City, State.Code, Zip.Code)) |>
  select(Street.Address, City, Address) |>
  head()
```

```
## # A tibble: 6 × 3
##   Street.Address       City       Address                              
##   &lt;chr&gt;          &lt;chr&gt;&lt;chr&gt;                                
## 1 ONE HOSPITAL ROAD    OAK BLUFFS ONE HOSPITAL ROAD OAK BLUFFS MA 2557 
## 2 2033 MAIN STREET     ATHOL      2033 MAIN STREET ATHOL MA 1331       
## 3 450 BROOKLINE AVE    BOSTON     450 BROOKLINE AVE BOSTON MA 2215     
## 4 145 WARD HILL AVENUE BRADFORD   145 WARD HILL AVENUE BRADFORD MA 1835
## 5 222 STATE STREET     LUDLOW     222 STATE STREET LUDLOW MA 1056      
## 6 1200 CENTRE STREET   ROSLINDALE 1200 CENTRE STREET ROSLINDALE MA 2127
```
---
      
# `mutate()`


```r
hospitals_ma |>
  mutate(Hospital.Name.Lower = tolower(Hospital.Name)) |>
  select(Hospital.Name, Hospital.Name.Lower) |>
  head()
```

```
## # A tibble: 6 × 2
##   Hospital.Name                       Hospital.Name.Lower                
##   &lt;chr&gt;                         &lt;chr&gt;                              
## 1 MARTHAS VINEYARD HOSPITAL           marthas vineyard hospital          
## 2 ATHOL MEMORIAL HOSPITAL             athol memorial hospital            
## 3 DANA-FARBER CANCER INSTITUTE        dana-farber cancer institute       
## 4 WHITTIER HOSPITAL-BRADFORD          whittier hospital-bradford         
## 5 HEALTHSOUTH REHABILITATION HOSPITAL healthsouth rehabilitation hospital
## 6 HEBREW REHABILITATION CENTER        hebrew rehabilitation center
```
---
      
# `mutate()`


```r
hospitals_ma |>
  mutate(Percentage.State.Beds = Number.of.Beds/sum(Number.of.Beds, na.rm = TRUE) * 100) |>
  select(Number.of.Beds, Percentage.State.Beds) |>
  head()
```

```
## # A tibble: 6 × 2
##   Number.of.Beds Percentage.State.Beds
##      &lt;int&gt;           &lt;dbl&gt;
## 1             25                 0.133
## 2             25                 0.133
## 3             30                 0.160
## 4             60                 0.319
## 5             53                 0.282
## 6            667                 3.55
```

---

# `group_by()` |> `mutate()`
      
&gt; How do you expect this to be different than `group_by()` |> `summarize()`?    
&gt; What do you expect this line of code to do?


```r
hospitals_ma |>
  group_by(City) |>
  mutate(Total.Beds.City = sum(Number.of.Beds, na.rm = TRUE))
```
      
---
      
# `group_by()` |> `mutate()`


```r
hospitals_ma |>
  group_by(City) |>
  mutate(Total.Beds.City = sum(Number.of.Beds, na.rm = TRUE)) |>
  arrange(desc(City)) |>
  select(City, Total.Beds.City) |>
  head()
```

```
## # A tibble: 6 × 3
## # Groups:   City [2]
##   City      Total.Beds.City
##   &lt;chr&gt;   &lt;int&gt;
## 1 WORCESTER            1373
## 2 WORCESTER            1373
## 3 WORCESTER            1373
## 4 WORCESTER            1373
## 5 WORCESTER            1373
## 6 WOBURN                210
```      

---
      
# `group_by()` |> `mutate()`


```r
hospitals_ma |>
  group_by(City) |>
  mutate(Total.Beds.City = sum(Number.of.Beds, na.rm = TRUE)) |>
  arrange(desc(Total.Beds.City)) |>
  select(City, Total.Beds.City) |>
  head()
```
      
```
## # A tibble: 6 × 3
## # Groups:   City [2]
##   City      Total.Beds.City
##   &lt;chr&gt;   &lt;int&gt;
## 1 BOSTON               4300
## 2 BOSTON               4300
## 3 BOSTON               4300
## 4 BOSTON               4300
## 5 BOSTON               4300
## 6 BOSTON               4300
```

