---
title: "Lec 05: In-class Exercise: Boxplots"
subtitle: "SDS 192: Introduction to Data Science"
author: |
   Shiya Cao
output: pdf_document
---

# Step 1: Load packages

```{r}
#Load packages
library(tidyverse)
library(spotifyr)
```

# Step 2: Create an Spotify Developer Account

Copy client id and secret from your previous in-class exercise into the chunk below to replace the current client id and secret, and then run the code chunk.

```{r}
id <- 'c283abfcc1c94026a6075358ab4ecfbe'
secret <- '3ee9ec4cd1cf433a8d3be549212f2a67'
Sys.setenv(SPOTIFY_CLIENT_ID = id)
Sys.setenv(SPOTIFY_CLIENT_SECRET = secret)
access_token <- get_spotify_access_token()
```

# Step 3: Get Song Features for Your Favorite Artist

Replace the text Janelle Monae below with the name of your favorite artist and then run the code chunk. 

```{r}
artist <- get_artist_audio_features(artist = "Janelle Monae") |>
  select(-c(album_images, artists, available_markets))
```

# Step 4: Create Grouped Boxplots Visualizing the Distribution of Values in a Song Feature of Your Choice, Grouped by Album Name

Refer to the [`ggplot()` cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf) to find the geom function you need to create this plot. 

Be sure to add labels to your plot. 

```{r}
# Fill your code here. 
```

# Step 5: Interpret Results

Turn to a neighbor and discuss the plots. 
* Check for outliers.
* Compare medians. 
* Compare the ranges.
* Compare the IQRs.
* Compare the symmetry. 

> Call us over if you are having trouble interpreting!


