---
title: "Lec 04: In-class Exercise: Frequency Plots and Facets"
subtitle: "SDS 192: Introduction to Data Science"
author: |
   Shiya Cao
output: pdf_document
---

# Today's Dataset

1. Spotify has an Application Programming Interface (API) that allows us to access data about music on the platform.
2. We can access data about specific songs, playlists, and artists.
3. Today we are going to access data about the tracks for a few different artists.
4. Variables include things such as acousticness, danceability, and speechiness, album information, and key.

# Step 1: Install and Load Packages

In your console install the Spotify R package:

`install.packages("spotifyr")`

Then load the packages for this lab by running the code below. 

```{r}
#Load packages
library(tidyverse)
library(spotifyr)
```

# Step 2: Create an Spotify Developer Account

1. Log-in to Spotify.com or create an account.
2. Go to [https://developer.spotify.com/dashboard](https://developer.spotify.com/dashboard)
3. Create an app named "SDS 192 Lec 04 In-class Exercise". You can indicate that this is "for frequency plots and facets".
4. Click "SHOW CLIENT SECRET".
5. Copy client id and secret into the chunk below to replace the current client id and secret, and then run the code chunk.

```{r}
id <- 'c283abfcc1c94026a6075358ab4ecfbe'
secret <- '3ee9ec4cd1cf433a8d3be549212f2a67'
Sys.setenv(SPOTIFY_CLIENT_ID = id)
Sys.setenv(SPOTIFY_CLIENT_SECRET = secret)
access_token <- get_spotify_access_token()
```

# Step 3: Get Song Features for Your Favorite Artist

Replace the text Janelle Monae below with the name of your favorite artist and then run the code chunk. 

```{r}
artist <- get_artist_audio_features(artist = "Janelle Monae") |>
  select(-c(album_images, artists, available_markets))
```

# Step 4: Create a Histogram Visualizing the Distribution of Values in a Song Feature of Your Choice

Be sure to add labels to your plot. 

```{r}
# Fill your code here. 
```

# Step 5: Facet Your Previous Plot by Album Name

Be sure to add labels to your plot. 

```{r}
# Fill your code here. 
```

Turn to a neighbor and discuss the plots. 
* What do the distributions indicate?
* What do you learn when comparing distributions across albums?

# Step 6: Create a Stacked, Dodged, or Filled Barplot Visualizing the Frequency of Songs in Each Album and Key Mode for This Artist

Be sure to add labels to your plot. 

```{r}
# Fill your code here. 
```

Turn to a neighbor and discuss the plots. 
* What do the frequencies indicate?
* What do you learn when comparing frequencies across albums?
